Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-17 tex/classical.tex
Reworded sec. 2.2

From the Boltzmann kinetic theory of gases, the probability for a particle to be in a given
state of position and velocity (\textit{aka} the distribution function) must satify
the transport equation (Eq.~(\ref{eq:transport}), below).
This equation is traditionally obtained by exploiting the volume-invariance of the $\mu$-space
(\textit{i.e.} the space formed by the direct product of position and velocity)
and considering that elastic collisions constitute the predominant mechanism
for exchanging energy among particles \cite{bib:huang}.
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(f_2'f_1'-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($f'_1$ and $f'_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

@@

%-------------------------
\subsection{H-theorem and the Maxwell-Boltzmann distribution}
The $H$-theorem can be proved using the $H$ functional defined by Boltzmann 
%
\begin{equation}\label{eq:hfunctional}
   H=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v,
\end{equation} 
%
where $f(\vec{v},t)$ is a homogeneous distribution function.
This $H$ functional describes a system with a global energy $E$, a global
number of free classical particles $N$, global volume $V$, and a temperature $T$.
Besides, the Boltzmann's $H$ functional correspond to a density of entropy
when the system arrives at the equilibrium state.

The $H$-theorem establishes that if $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and the molecular chaos hypothesis, then
$dH/dt\leq0$ and the system arrives to the equilibrium state when
$dH/dt=0$.
The theorem can be proved by applying the time derivative
to Eq. (\ref{eq:hfunctional}) and using the Boltzmann transport equation
(\ref{eq:transport}). As a consequence of this theorem, the distribution function
that describes the system in equilibrium corresponds to the Maxwell-Boltzmann
distribution. This distribution can be obtained by solving a sufficient
condition for $f$ to obtain the equilibrium distribution function $f_0$:
%
\begin{equation}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%

This condition can be obtained from the binary collision term of the Boltzmann
transport equation (\ref{eq:transport}).

The $H$-theorem fixes the time evolution direction of systems in nature, such
as the second law of thermodynamics does.\footnote{The molecular hypothesis is %
fundamental to determine the physical meaning of the $H$-theorem. For more %
information, see \cite{bib:huang}.}

>>>>

From the Boltzmann kinetic theory of gases, the probability for a particle to be in a given
state of position and velocity at a time $t$,
$f(\vec r,\vec v,t)$ (\textit{aka} the distribution function), must satify
the transport equation (Eq.~(\ref{eq:transport}), below).
This equation is traditionally obtained by exploiting the volume-invariance of the $\mu$-space
(\textit{i.e.} the space formed by the direct product of position and velocity)
and considering that elastic collisions constitute the predominant mechanism
for exchanging energy among particles \cite{bib:huang}.
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(f_2'f_1'-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($f'_1$ and $f'_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

@@

%-------------------------
\subsection{H-theorem and the Maxwell-Boltzmann distribution}

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the functional \cite{bib:tolman,bib:huang}
%
\begin{equation}\label{eq:hfunctional}
   H=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Here, $f(\vec{v},t)$ is a homogeneous distribution function, \textit{i.e.}
$f$ is assumed to be independent of $\vec r$.
The functional $H$, originally introduced by Boltzmann in 1872,
describes a diluted gas, at temperature $T$, with a total energy $E$, a total
number of free classical particles $N$, and total volume $V$.

Subsequently, the time-evolution of the gas towards equilibrium is obtained through the
$H$-theorem. This theorem establishes that if $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and the molecular chaos hypothesis is valid, then
$dH/dt\leq0$ in general, and if $dH/dt=0$, then the system is in the equilibrium state.
The $H$-theorem is proven by taking the time derivative
of Eq. (\ref{eq:hfunctional}) and using the Boltzmann transport equation
(\ref{eq:transport}) \cite{bib:huang}. The $H$-theorem fixes the direction of time-evolution
of systems in nature, which is consistent with the second law of thermodynamics; in fact,
$H$ can be associated with an entropy density.

On the other hand, considering a diluted gas in equilibrium (\textit{i.e.}
$(\partial f/\partial t)=0$),
with no applied external forces (\textit{i.e.} $f$ is independent of $\vec r$), 
it can be proven that the equilibrium distribution function obtained from
Eq.~(\ref{eq:transport}) is precisely the Maxwell-Boltzmann
distribution function. The proof of the above commences by finding the sufficient
condition for $f$ to render a null r.h.s. of Eq. (\ref{eq:transport}). Such an
$f$, which we denote here as $f_0$, must satisfy:
%
\begin{equation}\label{eq:neccondf0}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%
Subsequently, the Maxwell-Boltzmann distrubution function can be obtained by
taking the logarithm to Eq.~(\ref{eq:neccondf0}), and exploiting mechanical
conserved quantities (see \cite[ch. 4.2]{bib:huang}).

We emphasize that the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is surmised.
